The Construction of Set-Truncated Higher Inductive Types

### 2019-04-01 paper/paper.tex
Draft

\usepackage{cleveref}
\usepackage{entcsmacro}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{bussproofs}
\usepackage{stmaryrd}
\usepackage{cleveref}
\usepackage{listings}
\usepackage{coq}
\usepackage{color}
\usepackage{xcolor}

@@

% HoTT
\newcommand{\deprod}[3]{\prod(#1 : #2), #3} % dependent product
%\newcommand{\deprod}[3]{\prod_{#1 : #2} #3} % dependent product
\newcommand{\desum}[3]{\sum(#1 : #2), #3} % dependent sum
%\newcommand{\desum}[3]{\sum_{#1 : #2} #3} % dependent sum
\newcommand{\hset}{\type{hSet}} % sets
\newcommand{\0}{\textbf{0}} % empty type
\newcommand{\1}{\textbf{1}} % unit type
\newcommand{\TT}{\constructor{tt}} % element of unit type
\newcommand{\inl}{\term{inl}} %left inclusion
\newcommand{\inr}{\term{inr}} % right inclusion
\newcommand{\prl}{\term{pr}_1} % first projection
\newcommand{\prr}{\term{pr}_2} % second projection
\newcommand{\pairTm}[2]{(#1 , #2)} % pairing
\newcommand{\weq}{\simeq} % equivalence
\newcommand{\setquot}[2]{#1 / #2} % quotient type
\newcommand{\transport}[2]{\function{transport}^{#1} \> #2} % transport
\newcommand{\invP}[1]{(#1)^{-1}} % inverse of a path
\newcommand{\classof}{\constructor{class}} % class construcor
\newcommand{\classeq}{\constructor{classeq}} % path between classes
\newcommand{\isaprop}[1]{\function{isaprop}(#1)} % a type is a set
\newcommand{\isaset}[1]{\function{isaset}(#1)} % a type is a set
\newcommand{\trunct}[1]{\lVert #1 \rVert} % propositional truncation
\newcommand{\truncel}[1]{| #1 |} % constructor of truncation
\newcommand{\depeq}[3]{#2 =_{#1} #3} % dependent equality type
\newcommand{\Def}{:=} % symbol for definition

@@

% Adjunction
\newcommand{\quotF}{\mathcal{Q}}
\newcommand{\pathF}{\pi_0}
\newcommand{\quotPA}{\mathcal{Q}^{\category{PreAlg}}}
\newcommand{\pathPA}{\pi^{\category{PreAlg}_0}}
\newcommand{\quotAA}{\mathcal{Q}^{\category{Alg}}}
\newcommand{\pathAA}{\pi_0^{\category{Alg}}}

@@

\begin{example}
\label{ex:modsig}
Recall $\modZ$ from the introduction.
We represent it by the signature $\modsig$ defined as follows
\begin{itemize}
	\item $\pointc{\modsig} \Def \sumP{\I}{(\C \> \1)}$;
	\item $\pathI{\modsig} \Def \1$;
	\item $\pathA{\modsig} \Def \I$;
	\item $\pathlh{\modsig} \> j \Def \comp{\comp{\comp{\comp{\id{\I}}{\inle}}{\constr}}{\inle}}{\constr}$;
	\item $\pathrh{\modsig} \> j \Def \id{\I}$;
\end{itemize}
\end{example}

Intuitively, this signature represents a HIT $H$ with operation $H + \1 \rightarrow H$ representing the successor and zero.
Note that the endpoint $\comp{\inle}{\constr}$ takes the successor, so, the equation says that $S(S \> x) = x$ for all $x : X$.
Similarly, one can define signatures for groups or join semi-lattices.
Since signatures can depend on types, we can also define the propositional truncation.

\begin{example}
\label{ex:truncsig}
Let $A$ be a set.
We represent the truncation of a set by the following signature.
\begin{itemize}
	\item $\pointc{\trunc} \Def \C \> A$;
	\item $\pathI{\trunc} \Def \1$;
	\item $\pathA{\trunc} \Def \prodP{\I}{\I}$;
	\item $\pathlh{\trunc} \> j \Def \comp{\id{\prodP{\I}{\I}}}{\prle}$;
	\item $\pathrh{\trunc} \> j \Def \comp{\id{\prodP{\I}{\I}}}{\prre}$;
\end{itemize}
\end{example}
This signature represents a HIT $H$ with a map $A \rightarrow H$ and for which all inhabitants are equal.

@@

\begin{example}
\label{ex:commalg}
Recall the signature $\modsig$ from Example \ref{ex:modsig}.
A prealgebra of $\modsig$ consists of a set $X$ and a map $f : X + \1 \rightarrow X$.
We define $S_X \> x \Def f(\inl \> x)$ and $Z_X = f(\inr \> \TT)$.
An algebra of $\modsig$ consists of a prealgebra $X$ such that for all $x : X \times X$, we have $S_X(S_X \> x) = x$.
\end{example}

@@

\begin{lemma}
\label{lem:comm}
We have
\begin{itemize}
	\item $\pathF$ commutes with sums and products.
	\item $\quotF$ commutes with sums and products.
\end{itemize}
This gives rise to two natural isomorphisms $n_1 : \nattrans{\compf{\pathF}{\semPT{\pointc{\sig}}}}{\compf{\semP{\pointc{\sig}}}{\pathF}}$
\end{lemma}

Now we use Lemmata \ref{lem:prealgadj} and \ref{lem:comm} to lift both the quotient and the path setoid functors and we get
\[
\quotPA : \functor{\prealgst{\sig}}{\prealg{\sig}}
\quad \quad
\pathPA : \functor{\prealg{\sig}}{\prealgst{\sig}}.
\]

\begin{lemma}
We have an adjunction $\adj{\quotPA}{\pathPA}$.
\end{lemma}

@@

Now we would like to apply this proposition to the adjunction obtained from Lemma \ref{lem:prealgadj}.

@@

Now we can lift the quotient and path setoid.
We can verify the conditions and we get

To show this adjunction lifts to an adjunction between algebras, we need to show that they map algebras to algebras.
We only show this for the path setoid

\begin{lemma}
path setoid endpoint
\end{lemma}

\begin{lemma}
Path setoid factors through algebras
\end{lemma}

>>>>

\usepackage{cleveref}
\usepackage{entcsmacro}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{bussproofs}
\usepackage{stmaryrd}
\usepackage{cleveref}
\usepackage{listings}
\usepackage{coq}
\usepackage{color}
\usepackage{xcolor}
\usepackage[all]{xy}

@@

% HoTT
\newcommand{\deprod}[3]{\prod(#1 : #2), #3} % dependent product
%\newcommand{\deprod}[3]{\prod_{#1 : #2} #3} % dependent product
\newcommand{\desum}[3]{\sum(#1 : #2), #3} % dependent sum
%\newcommand{\desum}[3]{\sum_{#1 : #2} #3} % dependent sum
\newcommand{\hset}{\type{hSet}} % sets
\newcommand{\0}{\textbf{0}} % empty type
\newcommand{\unitt}{\textbf{1}} % unit type
\newcommand{\TT}{\constructor{tt}} % element of unit type
\newcommand{\inl}{\term{inl}} %left inclusion
\newcommand{\inr}{\term{inr}} % right inclusion
\newcommand{\prl}{\term{pr}_1} % first projection
\newcommand{\prr}{\term{pr}_2} % second projection
\newcommand{\pairTm}[2]{(#1 , #2)} % pairing
\newcommand{\weq}{\simeq} % equivalence
\newcommand{\setquot}[2]{#1 / #2} % quotient type
\newcommand{\transport}[2]{\function{transport}^{#1} \> #2} % transport
\newcommand{\invP}[1]{(#1)^{-1}} % inverse of a path
\newcommand{\classof}{\constructor{class}} % class construcor
\newcommand{\classeq}{\constructor{classeq}} % path between classes
\newcommand{\isaprop}[1]{\function{isaprop}(#1)} % a type is a set
\newcommand{\isaset}[1]{\function{isaset}(#1)} % a type is a set
\newcommand{\trunct}[1]{\lVert #1 \rVert} % propositional truncation
\newcommand{\truncel}[1]{| #1 |} % constructor of truncation
\newcommand{\depeq}[3]{#2 =_{#1} #3} % dependent equality type
\newcommand{\Def}{:=} % symbol for definition

@@

% Adjunction
\newcommand{\quotF}{\mathcal{Q}}
\newcommand{\pathF}{\pi_0}
\newcommand{\quotPA}{\mathcal{Q}^{\category{PreAlg}}}
\newcommand{\pathPA}{\pi^{\category{PreAlg}}_0}
\newcommand{\quotAA}{\mathcal{Q}^{\category{Alg}}}
\newcommand{\pathAA}{\pi_0^{\category{Alg}}}

@@

\begin{example}
\label{ex:modsig}
Recall $\modZ$ from the introduction.
We represent it by the signature $\modsig$ defined as follows
\begin{itemize}
	\item $\pointc{\modsig} \Def \sumP{\I}{(\C \> \unitt)}$;
	\item $\pathI{\modsig} \Def \unitt$;
	\item $\pathA{\modsig} \Def \I$;
	\item $\pathlh{\modsig} \> j \Def \comp{\comp{\comp{\comp{\id{\I}}{\inle}}{\constr}}{\inle}}{\constr}$;
	\item $\pathrh{\modsig} \> j \Def \id{\I}$;
\end{itemize}
\end{example}

Intuitively, this signature represents a HIT $H$ with operation $H + \unitt \rightarrow H$ representing the successor and zero.
Note that the endpoint $\comp{\inle}{\constr}$ takes the successor, so, the equation says that $S(S \> x) = x$ for all $x : X$.
Similarly, one can define signatures for groups or join semi-lattices.
Since signatures can depend on types, we can also define the propositional truncation.

\begin{example}
\label{ex:truncsig}
Let $A$ be a set.
We represent the truncation of a set by the following signature.
\begin{itemize}
	\item $\pointc{\trunc} \Def \C \> A$;
	\item $\pathI{\trunc} \Def \unitt$;
	\item $\pathA{\trunc} \Def \prodP{\I}{\I}$;
	\item $\pathlh{\trunc} \> j \Def \comp{\id{\prodP{\I}{\I}}}{\prle}$;
	\item $\pathrh{\trunc} \> j \Def \comp{\id{\prodP{\I}{\I}}}{\prre}$;
\end{itemize}
\end{example}
This signature represents a HIT $H$ with a map $A \rightarrow H$ and for which all inhabitants are equal.

@@

\begin{example}
\label{ex:commalg}
Recall the signature $\modsig$ from Example \ref{ex:modsig}.
A prealgebra of $\modsig$ consists of a set $X$ and a map $f : X + \unitt \rightarrow X$.
We define $S_X \> x \Def f(\inl \> x)$ and $Z_X = f(\inr \> \TT)$.
An algebra of $\modsig$ consists of a prealgebra $X$ such that for all $x : X \times X$, we have $S_X(S_X \> x) = x$.
\end{example}

@@

\begin{lemma}
\label{lem:comm}
We have
\begin{itemize}
	\item $\pathF$ commutes with sums and products.
	\item $\quotF$ commutes with sums and products.
\end{itemize}
This gives rise to two natural isomorphisms $n_1^P : \nattrans{\compf{\pathF}{\semPT{P}}}{\compf{\semP{P}}{\pathF}}$ and $n_2 : \nattrans{\compf{\quotF}{\semPT{P}}}{\compf{\semP{P}}{\quotF}}$.
\end{lemma}

Using Lemmata \ref{lem:prealgadj} and \ref{lem:comm}, we lift lift both the quotient and the path setoid functors to obtain.
\[
\quotPA : \functor{\prealgst{\sig}}{\prealg{\sig}}
\quad \quad
\pathPA : \functor{\prealg{\sig}}{\prealgst{\sig}}.
\]

\begin{proposition}
\label{lem:prealgadjconstr}
Given are categories $\mathcal{C}$ and $\mathcal{D}$, functors $A_1 : \functor{\mathcal{C}}{\mathcal{C}}$, $A_2 : \functor{\mathcal{D}}{\mathcal{D}}$, and $L : \functor{\mathcal{C}}{\mathcal{D}}$, and a natural transformation $n : \nattrans{\compf{L}{A_2}}{\compf{A_1}{L}}$.
In addition, given is a functor $R : \functor{\mathcal{D}}{\mathcal{C}}$ together with a natural transformation $m : \nattrans{\compf{R}{A_1}}{\compf{A_2}{L}}$ and an adjunction $\adj{L}{R}$ with unit $\eta$ and counit $\varepsilon$.
We also assume the following
\begin{itemize}
	\item For each $X : \mathcal{C}$, the following diagram commutes
	\[
	\xymatrix
	{	
		A_2 (L(R \> X)) \ar[dr]^{\epsilon \> X} \ar[d]_{\theta_1} \\
		R(L(A_2 \> X)) \ar[r]_{\epsilon(A_2 \> X)} & A_2 \> X
	}
	\]
	where the transformation $\theta_1$ is defined using whiskering, associativity of composition, and $n$ and $m$.
	\item For each $X : \mathcal{D}$, the following diagram commutes
	\[
	\xymatrix
	{
		A_1 \> X \ar[dr]_{\eta \> (A_1 \> X)} \ar[r]^{A_1 \> (\eta \> X)} & A_1(R(L \> X)) \ar[d]^{\theta_2}\\
		& R(L(A_1 \> X))
	}
	\]
	where the transformation $\theta_2$ is defined using whiskering, associativity of composition, and $n$ and $m$.
\end{itemize}
With all this, for each $X : \mathcal{D}$ the map $\eta \> X$ is an algebra homomorphism and for each $Y : \mathcal{C}$ the map $\epsilon \> X$ is an algebra homomorphism.
Furthermore, we have an adjunction $\adj{L^{\category{PreAlg}}}{R^{\category{PreAlg}}}$.
\end{proposition}

For the verification of the conditions of this proposition for the path setoid and quotient, we refer the reader to the formalization.
Then we can conclude

\begin{lemma}
\label{lem:adjprealg}
We have an adjunction $\adj{\quotPA}{\pathPA}$.
\end{lemma}

@@

Now we would like to apply this proposition to the adjunction obtained from Lemma \ref{lem:adjprealg}.
For that, we first need to calculate the action of the endpoints.
This is done by the following lemma.

\begin{lemma}
Given is an endpoint $e : \ep{A}{P}{Q}$.
For every set prealgebra $X$, we have the following equality
\[
\compm{\pathF \> (\semE{e} \> X)}{n_1^P \> X} = \compm{n_1^Q \> X}{\semET{e} \> (\pathPA \> X)}.
\]
For every setoid prealgebra $X$, we have the following equality.
\[
\compm{\quotF \> (\semET{e} \> X)}{n_2^P \> X} = \semE{e} \> (\quotPA \> X).
\]
\end{lemma}

Now we obtain the desired adjunction.

@@
